---
title: "Practical Machine Learning"
author: "Pradeep Senapati"
date: "April 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<font size="5"> **Introduction**</font>

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The project can be found on github: http://benakiva.github.io/practicalML/ and https://github.com/benakiva/practicalML

<font size="5"> **Data Source**</font>

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

<font size="5"> **Loading the Dataset**</font>

Files were downloaded from the URLs offline outside the program. Data from the downloaded files are loaded to dt_training and dt_testing data frames.

```{r}
library(plyr)
library(ggplot2)
```

```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(gbm)
library(gdata)
setwd ("C:/Pradeep/Data Science/Practical Machine Learning")
```
```{r}
# Load the training dataset
dt_training <- read.csv("C:/Pradeep/Data Science/Practical Machine Learning/pml-training.csv")
# Load the testing dataset
dt_testing <- read.csv("C:/Pradeep/Data Science/Practical Machine Learning/pml-testing.csv")
```

<font size="5"> **Cleaning Data**</font>

Removed all columns that contains NA and features that are not in the testing dataset. The features containing NA are the variance, mean and standard devition (SD) within each window for each feature. Since the testing dataset has no time-dependence, these values are useless and can be disregarded. Also removed the first 7 features since they are related to the time-series or are not numeric.

```{r}
features <- names(dt_testing[,colSums(is.na(dt_testing)) == 0])[8:59]

# Only use features used in testing cases.
dt_training <- dt_training[,c(features,"classe")]
dt_testing <- dt_testing[,c(features,"problem_id")]

dim(dt_training) 
dim(dt_testing)
```

<font size="5"> **Partitioning the Dataset**</font>

Data is split into a training data set (60% of the total cases) and a testing data set (40% of the total cases); the latter should not be confused with the data in the pml-testing.csv file). This will allow us to estimate the out of sample error of our predictor.

```{r}
inTrain <- createDataPartition(dt_training$classe, p=0.6, list=FALSE)
training <- dt_training[inTrain,]
testing <- dt_training[-inTrain,]

dim(training) 
dim(testing)
```

<font size="5"> **Building the Decision Tree Model**</font>

```{r}
set.seed(12345)
modFitDT <- rpart(classe ~ ., data = training, method="class", control = rpart.control(method = "cv", number = 10))
fancyRpartPlot(modFitDT)
```

<font size="5"> **Predicting with the Decision Tree Model**</font>

```{r}
library(e1071)
set.seed(12345)
prediction <- predict(modFitDT, testing, type = "class")
confusionMatrix(prediction, testing$classe)
```

<font size="5"> **Building the Random Forest Model**</font>

Using random forest, the out of sample error should be small. The error will be estimated using the 40% testing sample. We should expect an error estimate of < 3%.

```{r}
set.seed(12345)
modFitRF <- randomForest(classe ~ ., data = training, method = "rf", importance = T, trControl = trainControl(method = "cv", classProbs=TRUE,savePredictions=TRUE,allowParallel=TRUE, number = 10))
plot(modFitRF)
```

<font size="5"> **Building the Boosting Model**</font>

```{r}
modFitBoost <- train(classe ~ ., method = "gbm", data = training,
                    verbose = F,
                    trControl = trainControl(method = "cv", number = 10))

modFitBoost
```

```{r}
plot(modFitBoost)
```

<font size="5"> **Predicting with the Random Forest Model**</font>

```{r}
prediction <- predict(modFitRF, testing, type = "class")
confusionMatrix(prediction, testing$classe)
```

<font size="5"> **Predicting with the Boosting Model**</font>

```{r}
prediction <- predict(modFitBoost, testing)
confusionMatrix(prediction, testing$classe)
```

<font size="5"> **Predicting with the Test Data**</font>


<font size="4"> **Decision Tree Prediction**</font>

```{r}
predictionDT <- predict(modFitDT, dt_testing)
predictionDT
```

<font size="4"> **Random Forest Prediction**</font>

```{r}
predictionRF <- predict(modFitRF, dt_testing)
predictionRF
```

<font size="4"> **Boosting Prediction**</font>
```{r}
predictionBoost <- predict(modFitBoost, dt_testing)
predictionBoost
```

<font size="5"> **Conclusion**</font>

As can be seen from the confusion matrix the Random Forest model is very accurate, about 99%. Because of that we could expect nearly all of the submitted test cases to be correct. It turned out they were all correct.

Prepare the submission.
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionRF)
```

